---
title: "Baseline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Baseline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(baselinedetection)
```
# Introduction

This vignette will demonstrate how to use the core functions of the `yourpackagename` package. We will cover:

1. Importing spectral data from a CSV file.
2. Creating a `Spectra` object.
3. Performing baseline correction on the spectra.
4. Plotting the original and corrected spectra, along with the convex hull.


### Why Convex Hull and Baseline Correction?

When working with spectral data, it is common to encounter noise and background signals that can obscure the true features of the spectrum. To address this, we use **baseline correction**, a process that removes the unwanted background signal, allowing us to focus on the actual data of interest.

A common method for baseline correction is to use the **convex hull**. In simple terms, the convex hull is the smallest "envelope" that can completely enclose the data points of a spectrum. Imagine stretching a rubber band around the outermost points of the spectrum; the shape formed by the rubber band is analogous to the convex hull. By subtracting this envelope from the original spectrum, we effectively remove the background noise, resulting in a "flattened" spectrum that highlights the true peaks and valleys of the data.

In this vignette, we will show you how to perform these steps using the `baselinedetection`package.


# Importing Spectral Data

We start by importing spectral data from a CSV file using the `import_spectra` function. This function reads the data, checks that the required columns (`wavelength` and `intensity`) are present, and creates a `Spectra` object.

```{r, echo=TRUE}
library(baselinedetection)

# The CSV file has columns "wavelength" and "intensity"
file_path <- system.file("extdata", "test_spectra.csv", package = "baselinedetection")
spectra <- import_spectra(file_path)

print(spectra)
```

# Baseline Correction

Next, we use the baseline_correct_Spectra function to correct the baseline of the spectra. This function calculates the convex hull of the spectrum and subtracts it to correct the baseline.

```{r}
corrected_spectra <- baseline_correct_Spectra(spectra)

print(corrected_spectra)


```
# Plotting the spectra

Finally, we plot the original spectrum, the convex hull, and the baseline-corrected spectrum using the plot.Spectra function.

```{r}
hull_points <- convex_hull_Spectra(spectra) # Use thE Rcpp file to cacluate the hull points

# Plot the original, hull, and corrected spectra
plot(spectra, hull_points = hull_points, corrected_spectrum = corrected_spectra)

```

# Example

```{r}
 library(baselinedetection)
  wavelength <- seq(400, 700, length.out = 100)
  intensity <- sin(wavelength / 100) + rnorm(100, sd = 0.05)
  spectra <- as.Spectra(wavelength, intensity)
  corrected_spectrum <- baseline_correct_Spectra(spectra)
  hull_points <- convex_hull(spectra)
  plot(spectra)
  plot(spectra,hull_points=hull_points, corrected_spectrum = corrected_spectrum)
  plot_spectra_corrected(corrected_spectrum)

```
