---
title: "Baseline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Baseline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  
)
```

```{r}
library(baselinedetection)
```

# Introduction

This vignette will demonstrate how to use the core functions of the `baselinedetection` package. We will cover:

1. Importing spectral data from a CSV file.
2. Creating a `Spectra` object.
3. Performing baseline correction on the spectra.
4. Plotting the original and corrected spectra, along with the convex hull.



# 1) Background

## 1.1)  History of Baseline Detection

Baseline detection has been a fundamental aspect of spectral data analysis for decades, especially in fields like chemistry and physics, where spectroscopic data often include both signal and background noise. Originally, baseline detection was done manually or through simpler smoothing techniques. With advancements in computational methods, automated baseline correction techniques have emerged, allowing for more accurate and reproducible results. Early approaches involved simple manual subtraction of background signals, but as data complexity increased, automated methods were developed. Today, baseline correction techniques are a fundamental part of data preprocessing in spectroscopy, ensuring that analyses are accurate and reproducible.

## 1.2) Why is Baseline Correction Important?

In spectral analysis, baseline correction is essential because it removes unwanted background signals, which can obscure the true features of the data. This helps to accurately identify and quantify peaks, essential for applications such as material identification, chemical concentration analysis, and biomolecular characterization. Proper baseline correction also improves the reliability of any downstream analysis, such as peak fitting or machine learning applications.

## 1.3) Common Methods of Baseline Correction

There are several methods for baseline correction, each with unique advantages and limitations:

+ Polynomial Fitting: Fits a polynomial to the background noise. It’s simple but may not capture complex baselines.

+ Asymmetric Least Squares (ALS): A widely used method that penalizes deviations from smoothness, adjusting for asymmetry.

+ Convex Hull: Uses the smallest convex boundary around the data points, especially useful for data with sharp changes.


## 1.4 what is Convex Hull? 

The convex hull is the smallest convex boundary that can enclose all data points in the spectrum. Think of it as stretching a rubber band around the outermost points of the spectrum to create a smooth envelope. This method effectively handles varied spectral shapes and is robust against noise, making it ideal for baseline correction in spectroscopy.

The convex hull method subtracts this “envelope” from the original data, leaving a "flattened" spectrum where the true peaks and valleys are more prominent.

## 1.5) Montone chain algorithim 

Monotone Chain Algorithm for Convex Hull Calculation

Description: The Monotone Chain Algorithm, also known as Andrew’s Monotone Chain Algorithm, is a method for constructing the convex hull of a set of points. It sorts points by their x-coordinates (and by y-coordinates if x-coordinates are the same) and then constructs the convex hull in two steps: first by building the lower hull and then by building the upper hull.

Efficiency: The algorithm has a time complexity of O(nlogn), which comes from the initial sorting step. After sorting, the construction of the hull itself takes O(n), making the algorithm highly efficient and suitable for large datasets.

### Approach:

+ Sort Points: Points are sorted by x-coordinates. If two points have the same x-coordinate, they are sorted by y-coordinates.

+ Build Lower Hull: Starting from the leftmost point, the algorithm adds each point to the hull while ensuring only counterclockwise turns. It uses the orientation function to check the direction of the turn formed by the last two points in the hull and the current point.

+ Build Upper Hull: It then constructs the upper hull in a similar fashion, moving from rightmost to  leftmost.

+ Combine: The points from the lower and upper hulls are combined to form the convex hull.

### Orientation Function

The **orientation function** determines whether three points \( p \), \( q \), and \( r \) are collinear, or if they form a clockwise or counterclockwise turn. This is essential in the convex hull algorithm to ensure that only counterclockwise turns are included in the hull.

The orientation is calculated using the following formula:

\[
\text{val} = (q_y - p_y) \times (r_x - q_x) - (q_x - p_x) \times (r_y - q_y)
\]

where:
- If \( \text{val} = 0 \), the points \( p \), \( q \), and \( r \) are **collinear**.

- If \( \text{val} > 0 \), the sequence of points \( p \), \( q \), and \( r \) forms a **clockwise** turn.

- If \( \text{val} < 0 \), the sequence of points \( p \), \( q \), and \( r \) forms a **counterclockwise** turn.

This calculation helps in constructing the convex hull by ensuring that only points that form counterclockwise turns are added, thereby creating an "outer boundary" around the set of points.


# 2) Using the `baselinedetection` Package

## 2.1) Importing Spectral Data

We start by importing spectral data from a CSV file using the import_spectra function. This function reads the data, checks that the required columns (wavelength and intensity) are present, and creates a Spectra object.

```{r, echo=TRUE}
library(baselinedetection)

# The CSV file has columns "wavelength" and "intensity"
file_path <- system.file("extdata", "test_spectra.csv", package = "baselinedetection")
spectra <- import_spectra(file_path)

print(spectra)
```


## 2.2) Convex Hull Calculation

We first calculate the convex hull using the convex_hull_Spectra function, which identifies the points that form the baseline envelope.

```{r}
hull_points <- convex_hull_Spectra(spectra)
print(hull_points)

```

## 2.3) Performing Baseline Correction Using Convex Hull

The baseline_correct_Spectra function uses the convex hull to correct the baseline by subtracting the hull envelope from the original spectrum. Over here we use the interpolation and the approx function from base R.

```{r}
corrected_spectra <- baseline_correct_Spectra(spectra)
print(corrected_spectra)

```

## 2.4) Plotting the Spectra

The plot.Spectra function visualizes the original spectrum, convex hull, and corrected spectrum.

```{r plot-spectra, fig.width=7, fig.height=5}
hull_points <- convex_hull_Spectra(spectra) # Use thE Rcpp file to cacluate the hull points

# Plot the original, hull, and corrected spectra
plot(spectra, hull_points = hull_points, corrected_spectrum = corrected_spectra)

```

## 2.5) Plotting Only the Corrected Spectrum
You can also plot just the baseline-corrected spectrum for a cleaner view of the corrected data.

```{r  corrected plot-spectra, fig.width=7, fig.height=5}
plot_spectra_corrected(corrected_spectrum = corrected_spectra)
```

# Conclusion

This vignette demonstrates the core functions of the baselinedetection package, illustrating how to create, import, and plot spectral data, calculate the convex hull, and correct the baseline for clearer analysis.
